---
title: "Introduction to comexr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to comexr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The **comexr** package provides a complete R interface to the
[ComexStat API](https://comexstat.mdic.gov.br/) from the Brazilian
Ministry of Development, Industry, Trade and Services (MDIC). It gives
you programmatic access to:

- **General trade data** (1997--present): exports and imports with 22
  detail/grouping fields and 6 metrics.
- **City-level trade data**: trade statistics broken down by municipality.
- **Historical data** (1989--1996): legacy trade records using NBM codes.
- **Auxiliary tables**: countries, economic blocs, states, cities,
  transport modes, customs units, NCM/NBM/HS codes, and CGCE/SITC/ISIC
  classifications.

Only two dependencies: `httr2` and `cli`.

## Installation

```{r install}
# From GitHub
remotes::install_github("StrategicProjects/comexr")
```

```{r setup}
library(comexr)
```

## Quick Start

### Exports by country

```{r exports}
exports <- comex_export(
  start_period = "2024-01",
  end_period   = "2024-01",
  details      = "country"
)
head(exports)
```

By default, `month_detail = TRUE` so results include a month column.

### Imports with CIF

```{r imports}
imports <- comex_import(
  start_period = "2024-01",
  end_period   = "2024-12",
  details      = "country",
  metric_cif   = TRUE
)
```

### Filtered query

```{r filtered}
# Exports to China (id=160), grouped by HS4
china_hs4 <- comex_export(
  start_period = "2024-01",
  end_period   = "2024-12",
  details      = c("country", "hs4"),
  filters      = list(country = 160)
)
```

## Three Query Endpoints

| Function | Endpoint | Period | Detail names |
|----------|----------|--------|--------------|
| `comex_query()` / `comex_export()` / `comex_import()` | POST /general | 1997--present | NCM, HS, CGCE, SITC, ISIC, ... |
| `comex_query_city()` | POST /cities | 1997--present | heading, chapter, section |
| `comex_historical()` | POST /historical-data/ | 1989--1996 | country, state, nbm |

## General Query Details

These details work with `comex_query()`, `comex_export()`, and
`comex_import()`:

| Category | Details |
|----------|---------|
| **Geographic** | `country`, `bloc`, `state`, `transport_mode`, `customs_unit` |
| **Products** | `ncm`, `hs6`, `hs4`, `hs2`, `section` |
| **CGCE (BEC)** | `cgce_n1`, `cgce_n2`, `cgce_n3` |
| **SITC** | `sitc_section`, `sitc_chapter`, `sitc_position`, `sitc_subposition`, `sitc_item` |
| **ISIC** | `isic_section`, `isic_division`, `isic_group`, `isic_class` |
| **Other** | `company_size` |

Use `comex_details("general")` to confirm at runtime.

## General Query Metrics

| Parameter | Description | Available for |
|-----------|-------------|---------------|
| `metric_fob` | FOB value (US$) | export + import |
| `metric_kg` | Net weight (kg) | export + import |
| `metric_statistic` | Statistical quantity | export + import |
| `metric_freight` | Freight (US$) | import only |
| `metric_insurance` | Insurance (US$) | import only |
| `metric_cif` | CIF value (US$) | import only |

## City-Level Queries

The city endpoint uses **different detail names** from the general endpoint:

```{r city}
# Exports from Pernambuco by country
pe_exports <- comex_query_city(
  flow         = "export",
  start_period = "2024-01",
  end_period   = "2024-12",
  details      = c("state", "country"),
  filters      = list(state = 26)
)

# By product heading (SH4 equivalent)
pe_products <- comex_query_city(
  flow         = "export",
  start_period = "2024-01",
  end_period   = "2024-06",
  details      = c("state", "heading"),
  filters      = list(state = 26)
)
```

**Important differences from general queries:**

- Product details: use `heading` (SH4), `chapter` (SH2), `section` ---
  NOT `hs4`/`hs2`/`ncm`.
- Additional detail: `economic_block` (or `economicBlock`).
- Only FOB, KG, and statistical quantity metrics are available.
- Use `comex_details("city")` to see available fields.

## Historical Queries (1989--1996)

```{r historical}
hist_data <- comex_historical(
  flow         = "export",
  start_period = "1995-01",
  end_period   = "1996-12",
  details      = "country"
)
```

**Important differences:**

- Available details: only `country`, `state`, and `nbm`.
- Available metrics: only FOB and KG (no CIF, freight, insurance, or
  statistical quantity).
- Product classification uses NBM, not NCM.
- Use `comex_details("historical")` to confirm.

## Auxiliary Tables

### Geography

```{r tables_geo}
countries <- comex_countries()
comex_country_detail(105)   # Brazil

states <- comex_states()
comex_state_detail(26)      # Pernambuco

cities <- comex_cities()
comex_city_detail(2611606)  # Recife

blocs <- comex_blocs(language = "en")

transport <- comex_transport_modes()
comex_transport_mode_detail("01")  # Maritime

customs <- comex_customs_units()
```

### Products

```{r tables_products}
ncm <- comex_ncm(language = "en")
comex_ncm_detail("12019000")       # Soybeans

nbm <- comex_nbm()                 # Historical codes
comex_nbm_detail("2924101100")

hs <- comex_hs(language = "en")    # Harmonized System
```

### Classifications

```{r tables_class}
cgce <- comex_cgce(language = "en")
sitc <- comex_sitc(language = "en")
isic <- comex_isic(language = "en")
```

## Discovering Filter Values

Not all filters have a values endpoint. Use `comex_filter_values()` with
filters that are supported:

```{r filter_values}
# General: country, economicBlock, state, urf, ncm, section
comex_filter_values("country")
comex_filter_values("section")

# City: country, economicBlock, state, city, heading, chapter, section
comex_filter_values("heading", type = "city")

# Historical: country, state
comex_filter_values("country", type = "historical")
```

## SSL Certificate Issues

The ComexStat API uses the ICP-Brasil certificate chain, which may not
be recognized on all systems. The package handles this automatically:
on the first SSL failure it retries without verification and issues a
warning. To suppress the warning permanently:

```{r ssl}
options(comex.ssl_verifypeer = FALSE)
```

## Complete Example

```{r complete_example}
library(comexr)

# Step 1: Find country code
countries <- comex_countries()
countries[grepl("China", countries$text, ignore.case = TRUE), ]
# id = 160

# Step 2: Query soybean exports to China
soy_china <- comex_export(
  start_period = "2024-01",
  end_period   = "2024-12",
  details      = c("country", "hs4"),
  filters      = list(country = 160),
  month_detail = TRUE
)

# Step 3: Analyze
head(soy_china)
```

## References

- [ComexStat](https://comexstat.mdic.gov.br/) --- Brazilian foreign trade statistics
- [ComexStat API](https://api-comexstat.mdic.gov.br/docs) --- Official API documentation
- [MDIC](https://www.gov.br/mdic/) --- Ministry of Development, Industry, Trade and Services
